---
title: "Reproducible Research Course Project 1"
author: "Liliana Grajales"
date: "April 3, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Course Project 1

This document includes the code and graphs for Project 1 of the Reproducible Reseach Coursera Course. The data for this project is located at  <https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip>.

The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as )
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV)  and there are a total of 17,568 observations in this dataset.

```{r data}
#Libraries
library(dplyr)
library(base64)
library(ggplot2)
library(chron)

#Setting working directory path (when working from computer)
#setwd(~/repdata-data-activity")

#*****************************
# Reading the table and separating the string of data into columns using sep=";"
#****************************
activity_all <- read.csv("activity.csv", sep=",", colClasses ="character")
#Ommiting NAs
activity <-na.omit(activity_all)
#*****************************
# Plotting the data
#****************************
#First column date format description
day <- as.Date(activity[,2], "%Y-%m-%d")
steps <- as.numeric(activity[,1])
interval <-as.numeric(activity[,3])
steps_day <- data.frame(day,steps)
```

## What is mean total number of steps taken per day?

In Plot #1 below each circle represents the number of steps taken in intervals of 5 minutes each day
For more detail time series plot at each 5-minute interval see Plot # 5

```{r steps, echo=FALSE}
plot(day, steps, type ="p", xlab="date", ylab= "Steps", main="Number of Steps Taken each day (10/2-11/11), circle = steps in 5-min")
```

```{r stats}
#****************************
#Stats per day with NAs remmoved
#************************
sum_d <-with(steps_day, tapply(steps, day, sum))
mean_d <-with(steps_day, tapply(steps, day, mean))
std_d <-with(steps_day, tapply(steps, day, sd))
sqrt_dim_steps <-with(steps_day, tapply(steps, day, length))
sem_d<-std_d/sqrt(sqrt_dim_steps)
median_d <-with(steps_day, tapply(steps,day,median))
 
```

Plot #2 shows a Histogram of the total number of stepst taken each day.

```{r histogram, echo=FALSE}
hist(sum_d, xlab= "steps", main="Hystogram of total sum of steps per day without NAs (10/2-11/11)", col="red", ylim=c(0,40))
```

Plot #3 shows the mean of the total number of steps taken per day
```{r mean, echo=FALSE}
plot(unique(day), mean_d, type ="p", xlab="date", ylab= "Steps Mean", main="Mean Value of Steps Taken Per Day")
```

Plot #4 shows the median of the total number of steps taken per day
```{r median, echo=FALSE}
plot(unique(day), median_d, type ="p", xlab="date", ylab= "Steps Median", main="Median Value of Steps Taken Per Day")
```

## What is the average daily activity pattern?

Calculating the time series plot of the 5-minute interval and the average number of steps taken across all days 
```{r interval}
#***************************
#First two digit from left to right = Hour
#Last Two digits from left to right = Min
hour_d<-as.character(interval %/% 10^2) #First and second digit (2355= 23hour, 55 min)
minute_d<-as.character(interval %% 10^2) #last two digits
H_M_d <-paste(hour_d,minute_d, sep=":")
day_Hour_Min <-as.POSIXct(paste(day,H_M_d), format="%Y-%m-%d %H:%M")
all_steps_mean <-mean(steps)
```

Plot #5 shows a time series plot of the 5-minute interval and the average number of steps taken across all days (blue line)
```{r series, echo=FALSE}
plot(day_Hour_Min, steps, type ="p", pch=".", xlab="Interval", ylab= "Steps", main="Steps Taken During Each Interval Over Time")
#abline(day_Hour_Min,all_steps_mean)
abline(all_steps_mean,0, col="blue") # abline(intercept,slope)
```

Calculating Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r maxint }
max_steps <-max(steps)
five_minute_interval <- data.frame(interval, steps)
five_minute_interval_max <- subset(five_minute_interval, five_minute_interval$steps==max_steps)

```

The maximum number of steps done in a 5-minute interval are:
```{r maxint1}
max_steps
```

The 5-minute interval with the maximum number of steps is:

```{r maxint2}
five_minute_interval_max$interval
```

## Imputing missing values

The following code describes how to modify the missing step values or NAs

```{r Imputting}
#*******************************
#Imputting missing values
#************************
number_NAs <- dim(activity_all)-dim(activity)
#Missing values number
number_NAs[1]
#****************************
date_m<- as.Date(activity_all[,2], "%Y-%m-%d")
steps_m <- as.numeric(activity_all[,1])
interval_m <-as.numeric(activity_all[,3])
activity_modified <-data.frame(date_m,steps_m,interval_m)
#Below is the code to modify all NAs with the mean of all the steps
activity_modified$steps_m[which(is.na(activity_modified$steps_m))]<- all_steps_mean
#****************************
#Stats per day with NAs replaced by overall steps mean
#*********************************
sum_m <-with(activity_modified, tapply(steps_m, date_m, sum))
mean_m <-with(activity_modified, tapply(steps_m, date_m, mean))
std_m <-with(activity_modified, tapply(steps_m, date_m, sd))
sqrt_dim_steps_m <-with(activity_modified, tapply(steps_m, date_m, length))
sem_m<-std_m/sqrt(sqrt_dim_steps_m)
median_m <-with(activity_modified, tapply(steps_m,date_m,median))
plot(unique(activity_modified$date_m), mean_m, type ="p", xlab="date", ylab= "Steps Mean Modified", main="Mean Value of Steps Taken Per Day, NAs=Average")
plot(unique(activity_modified$date_m), median_m, type ="p", xlab="date", ylab= "Steps Median Modified", main="Mean Value of Steps Taken Per Day, NAs=Average")
stats_day_m <- data.frame(date1_m=unique(activity_modified$date_m),sum_m,mean_m,median_m,std_m,sem_m,sqrt_dim_steps_m)
```

Ploting the new histogram where the "NAs"" where modified by the overall average number of steps 

```{r hist2, echo=FALSE}
# Plotting the Hystogram
hist(sum_m, xlab= "steps modified", main="Hystogram of total sum of steps per day with Modified NAs (10/2-11/11)", col="red", ylim=c(0,40))
#****************************
```

## Are there diferences in activity patterns between weekdays and weekends?

Calculating Steps done during the weekdays and the Weekends

```{r week}
#****************************
#Separating weekend data from week data
#**********************************
hour_m<-as.character(interval_m %/% 10^2) #First and second digit (2355= 23hour, 55 min)
minute_m<-as.character(interval_m %% 10^2) #last two digits
H_M_m <-paste(hour_m,minute_m, sep=":")
day_Hour_Min_m <-as.POSIXct(paste(date_m,H_M_m), format="%Y-%m-%d %H:%M")
activity_modified$date_time <-day_Hour_Min_m 

activity_modified_wkend <-subset(activity_modified,is.weekend(activity_modified$date_m))
activity_modified_wk <-subset(activity_modified,!is.weekend(activity_modified$date_m))
all_steps_mean_wkend <-mean(activity_modified_wkend$steps_m)
all_steps_mean_wk <-mean(activity_modified_wk$steps_m)
```

Panel plot comparing the average number og stapes taken per 5-min interval across weekdays and weekend

```{r weekplt, echo=FALSE}
par(mfrow=c(1,2), margin(2,2,2,2))
plot(activity_modified_wk$date_time, activity_modified_wk$steps_m, pch=".", xlab="Week Days", ylab= "Steps with NAs replaced", main=" Steps Per Weekdays Mean=Red" )
abline(all_steps_mean_wk,0,col="red") # abline(intercept,slope)
plot(activity_modified_wkend$date_time, activity_modified_wkend$steps_m, pch=".",xlab="Weekend Days", ylab= "Steps with NAs replaced", main=" Steps Per Weekends Mean=Red")
abline(all_steps_mean_wkend,0,col="red") # abline(intercept,slope)
dev.off() #Closing png device
```

